---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error   = FALSE,
  message = FALSE,
  warning = FALSE,
  echo    = TRUE
)
```

# IdentifyingConservationGaps

<!-- badges: start -->

<!-- badges: end -->

This repository documents the data and R code used to reproduce the terrestrial analyses and **Fig. 1** in the manuscript
*Identifying Conservation Gaps: A Framework for Evaluating National Contributions to the 30x30 Target*.

Fig. 1 shows how Denmark’s terrestrial areas contribute to the **30% land target**, distinguishing between:

* areas that fully qualify as contributing protected areas,
* areas requiring individual assessment,
* areas with insufficient legal protection,
* areas compromised by infrastructure, forestry, or agriculture, and
* unprotected land.

This README focuses on the spatial data processing steps used to derive the terrestrial map and summary statistics for Fig. 1.

## Software and packages

All spatial calculations were done in R using the [`terra`](https://cran.r-project.org/package=terra) package. Visualisations were generated with [`ggplot2`](https://cran.r-project.org/package=ggplot2) and [`tidyterra`](https://cran.r-project.org/package=tidyterra). The terrestrial outline of Denmark was retrieved using the [`geodata`](https://cran.r-project.org/package=geodata) package.

```{r LoadPackages}
library(geodata)
library(ggplot2)
library(terra)
library(tidyterra)

```

## Spatial template and national boundary

All rasters used in the analysis are aligned to a common template defining the extent, resolution and coordinate reference system.

```{r Template}
Template <- terra::rast("Data/Template.tif")

DK <- geodata::gadm(country = "Denmark", level = 0, path = getwd(), version = "4.0") |> 
  terra::project(terra::crs(Template))
```

We then calculate the terrestrial area of Denmark in square kilometres:

```{r AreaKM}
Area_DK <- terra::expanse(DK, unit = "km")
```


Which is `r prettyNum((round(Area_DK, 2)), big.mark = ",")` square kilometers.


## Protected areas contributing to the 30% target

In the manuscript, areas that **fully contribute** to the 30% land target are those protection schemes that meet all criteria C1–C5 for protected areas (Tables 1 and 5 in the manuscript). On land, these are:

* **Approved Nature National Parks**
* **Areas owned by private nature foundations**
* **State-owned unmanaged / untouched forests**

Below we show how these layers are read, harmonised and combined into a single raster representing all areas that contribute to the 30% terrestrial target.

To track overlaps between protection schemes, each scheme is encoded as a binary layer with a **distinct base value** (1, 10, 100). When these layers are summed, each combination of schemes yields a unique value (e.g. 1 = Nature National Park only; 11 = Nature National Park + private foundation; 111 = all three schemes).

### Approved Nature National Parks

We read the raster of approved Nature National Parks and convert it to a binary layer where
0 = not an approved Nature National Park,
1 = approved Nature National Park.

```{r readInNatureNationalParks}
NNP_approved <- as.numeric(terra::rast("Data/Rast_Nature_National_Parks_Croped.tif")) + 1
NNP_approved <- terra::ifel(is.na(NNP_approved), 0, NNP_approved)
```

### Private nature foundations

Similarly, we read the raster of areas owned by private nature foundations (`Fondsejede`) and encode it as

0 = not foundation-owned,
10 = foundation-owned.

```{r readInPrivateFundations}
Fondsejede <- as.numeric(terra::rast("Data/Rast_Fondsejede_Croped.tif")) + 10
Fondsejede <- terra::ifel(is.na(Fondsejede), 0, Fondsejede)
```

### Unmanaged / untouched forests

We then read the raster of state-owned unmanaged / untouched forests (`UroertSkov_NST`) and encode it as

0 = not unmanaged / untouched forest,
100 = unmanaged / untouched forest.


```{r readInUntouchedForests}
UroertSkov_NST <- as.numeric(terra::rast("Data/Rast_Urort_Skov_Croped.tif")) + 99
UroertSkov_NST <- terra::ifel(is.na(UroertSkov_NST), 0, UroertSkov_NST)
```

### Generate composite layer of contributing protected areas

We now add the three layers to obtain a **composite raster** `PA30`, where the value of each pixel uniquely identifies which combination of schemes covers that pixel.

```{r CombineThem}
PA30 <- NNP_approved + Fondsejede + UroertSkov_NST

cls <- data.frame(
  ID    = c(1,   10,          11,             100,            101,            110,                  111),
  class = c("NNP_approved",
            "Fondsejede",
            "NNP_Fondsejede",
            "UroertSkov_NST",
            "NNP_Uroert",
            "Fondsejede_Uroert",
            "NNP_Fondsejede_Uroert")
)

PA30 <- terra::ifel(PA30 == 0, NA, PA30)
```

We then remove built-up areas and mask the result to the terrestrial area of Denmark.

Here, `BuildUp` is a raster where
0 = built-up (infrastructure, urban areas),
1 = other land.
This raster is prepared in the data pre-processing steps (not shown here).

```{r CorrectBuildup}
BuildUp <- terra::rast("Data/Rast_BuildUp_Croped.tif")
PA30 <- terra::ifel(BuildUp == 0, NA, PA30)

levels(PA30) <- cls

PA30 <- terra::mask(PA30, DK)
```

Finally, we save the composite layer as a Cloud Optimised GeoTIFF:

```{r SavePA30}
SpeciesPoolR::write_cog(PA30, "FinalLayers/PA30.tif")
```

This layer corresponds to the **turquoise category** (“Protected areas that fully meet the criteria”) in Fig. 1 of the manuscript.


## Production Areas

## Insuffient legal protection

## Requires individual assesment
