---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error   = FALSE,
  message = FALSE,
  warning = FALSE,
  echo    = TRUE
)
```

# IdentifyingConservationGaps

<!-- badges: start -->

<!-- badges: end -->

This repository documents the data and R code used to reproduce the terrestrial analyses and **Fig. 1** in the manuscript
*Identifying Conservation Gaps: A Framework for Evaluating National Contributions to the 30x30 Target*.

Fig. 1 shows how Denmark’s terrestrial areas contribute to the **30% land target**, distinguishing between:

* areas that fully qualify as contributing protected areas,
* areas requiring individual assessment,
* areas with insufficient legal protection,
* areas compromised by infrastructure, forestry, or agriculture, and
* unprotected land.

This README focuses on the spatial data processing steps used to derive the terrestrial map and summary statistics for Fig. 1.

## Software, packages and helper functions

All spatial analyses were carried out in R using the following core packages:

- [`terra`](https://cran.r-project.org/package=terra) for raster and vector processing,  
- [`ggplot2`](https://cran.r-project.org/package=ggplot2) and [`tidyterra`](https://cran.r-project.org/package=tidyterra) for visualisation, and  
- [`geodata`](https://cran.r-project.org/package=geodata) to obtain the national boundary of Denmark.

We also use [`stringr`](https://cran.r-project.org/package=stringr) for simple text wrapping in legend labels.

In addition, we define two small helper functions:

- `ToOne()` to read a raster from the `Data/` folder and convert all non-`NA` values to 1 (presence/absence), and  
- `write_cog()` to save a `SpatRaster` as a Cloud Optimized GeoTIFF (COG) using GDAL options for compression.

```{r LoadPackages}
library(geodata)
library(ggplot2)
library(terra)
library(tidyterra)
library(stringr)   # for stringr::str_wrap used below
```

```{r helpers}
# Helper to read a raster from the Data/ folder and convert all non-NA values to 1
ToOne <- function(x) {
  r <- terra::rast(file.path("Data", x))
  r <- as.numeric(r)

  r <- terra::ifel(!is.na(r), 1, 0)

  return(r)
}

# Helper to write a SpatRaster as a Cloud Optimized GeoTIFF (COG)
# COGs are geospatial files optimised for efficient cloud storage and retrieval.
write_cog <- function(SpatRaster, Name) {
  terra::writeRaster(
    x        = SpatRaster,
    filename = Name,
    overwrite = TRUE,
    gdal      = c("COMPRESS=DEFLATE", "TFW=YES", "of=COG")
  )
}
```

## Spatial template and national boundary

All rasters used in the analysis are aligned to a common template defining the extent, resolution and coordinate reference system.

```{r Template}
Template <- terra::rast("Data/Template.tif")

DK <- geodata::gadm(country = "Denmark", level = 0, path = getwd(), version = "4.0") |> 
  terra::project(terra::crs(Template))
```

We then calculate the terrestrial area of Denmark in square kilometres:

```{r AreaKM}
Area_DK <- terra::expanse(DK, unit = "km")
```


Which is `r prettyNum((round(Area_DK, 2)), big.mark = ",")` square kilometers.

## Build up

In many of the layers we need to correct using buildup, which was extracted from the landcover layer in [basemap](envs.au.dk/en/research-areas/society-environment-and-resources/land-use-and-gis/basemap)

Here, `BuildUp` is a raster where
0 = built-up (infrastructure, urban areas),
1 = other land.
This raster is prepared in the data pre-processing steps (not shown here).

```{r readinBuildup}
BuildUp <- terra::rast("Data/Rast_BuildUp_Croped.tif")
```



## Protected areas contributing to the 30% target

In the manuscript, areas that **fully contribute** to the 30% land target are those protection schemes that meet all criteria C1–C5 for protected areas (Tables 1 and 5 in the manuscript). On land, these are:

* **Approved Nature National Parks**
* **Areas owned by private nature foundations**
* **State-owned unmanaged / untouched forests**

Below we show how these layers are read, harmonised and combined into a single raster representing all areas that contribute to the 30% terrestrial target.

To track overlaps between protection schemes, each scheme is encoded as a binary layer with a **distinct base value** (1, 10, 100). When these layers are summed, each combination of schemes yields a unique value (e.g. 1 = Nature National Park only; 11 = Nature National Park + private foundation; 111 = all three schemes).

### Approved Nature National Parks

We read the raster of approved Nature National Parks and convert it to a binary layer where
0 = not an approved Nature National Park,
1 = approved Nature National Park.

```{r readInNatureNationalParks, eval=FALSE}
NNP_approved <- as.numeric(terra::rast("Data/Rast_Nature_National_Parks_Croped.tif")) + 1
NNP_approved <- terra::ifel(is.na(NNP_approved), 0, NNP_approved)
```

### Private nature foundations

Similarly, we read the raster of areas owned by private nature foundations (`Fondsejede`) and encode it as

0 = not foundation-owned,
10 = foundation-owned.

```{r readInPrivateFundations, eval=FALSE}
Fondsejede <- as.numeric(terra::rast("Data/Rast_Fondsejede_Croped.tif")) + 10
Fondsejede <- terra::ifel(is.na(Fondsejede), 0, Fondsejede)
```

### Unmanaged / untouched forests

We then read the raster of state-owned unmanaged / untouched forests (`UroertSkov_NST`) and encode it as

0 = not unmanaged / untouched forest,
100 = unmanaged / untouched forest.


```{r readInUntouchedForests, eval=FALSE}
UroertSkov_NST <- as.numeric(terra::rast("Data/Rast_Urort_Skov_Croped.tif")) + 99
UroertSkov_NST <- terra::ifel(is.na(UroertSkov_NST), 0, UroertSkov_NST)
```

### Generate composite layer of contributing protected areas

We now add the three layers to obtain a **composite raster** `PA30`, where the value of each pixel uniquely identifies which combination of schemes covers that pixel.

```{r CombineThem, eval=FALSE}
PA30 <- NNP_approved + Fondsejede + UroertSkov_NST

cls <- data.frame(
  ID    = c(1,   10,          11,             100,            101,            110,                  111),
  class = c("NNP_approved",
            "Fondsejede",
            "NNP_Fondsejede",
            "UroertSkov_NST",
            "NNP_Uroert",
            "Fondsejede_Uroert",
            "NNP_Fondsejede_Uroert")
)

PA30 <- terra::ifel(PA30 == 0, NA, PA30)
```

We then remove built-up areas and mask the result to the terrestrial area of Denmark.


```{r CorrectBuildup, eval=FALSE}
PA30 <- terra::ifel(BuildUp == 0, NA, PA30)

levels(PA30) <- cls

PA30 <- terra::mask(PA30, DK)
```

Finally, we save the composite layer as a Cloud Optimised GeoTIFF:

```{r SavePA30, eval=FALSE}
write_cog(PA30, "FinalLayers/PA30.tif")
```

This layer corresponds to the **turquoise category** (“Protected areas that fully meet the criteria”) in Fig. 1 of the manuscript.

## Production areas (areas compromising biodiversity)

In Fig. 1, areas where biodiversity is directly compromised by human land uses (e.g. intensive agriculture, production forestry, infrastructure) are shown in **orange** (“Areas compromised by infrastructure, forestry, or agriculture”).

These areas are identified by combining:

* agricultural land (fields in rotation and permanent grasslands not covered by Article 3 protection),
* production forest (tree-covered areas not designated as unmanaged forest or Article 3 areas), and
* infrastructure (roads, buildings and other built-up areas).

In the analysis pipeline, these inputs are assembled into a raster layer where pixels used primarily for these land uses are flagged as **compromised** and subsequently excluded from contributing to the 30% target. The resulting raster is then used to derive the orange category in Fig. 1.

### Reading and combining the layers

We use a raster (`Subclasses`) which classifies the areas that are within at least one protection scheme, that includes, among others, fields in rotation (`"INT_AGG"`), grasslands not covered by Article 3 protection (`"PGR_out_of_P3"`), and production forest (`"Drevet Skov"`). We merge these classes into a single layer of production areas.


```{r productionForest, eval=FALSE}
Subclasses<- rast("Data/Subclasses.tif")

ProductionAreas <- Template

ProductionAreas <- terra::ifel(Subclasses %in% c("Drevet Skov", "INT_AGG", "PGR_out_of_P3"), 1, 0)
```


We then add built-up areas (from `BuildUp`, read above). Pixels that are both within the `Subclasses` (this means that is part of some protection scheme) raster and classified as built-up (`BuildUp == 0`) are also flagged as production / compromised areas.

```{r productionForestBuildup, eval=FALSE}
ProductionAreas <- terra::ifel(!is.na(Subclasses) & BuildUp == 0, 1, ProductionAreas)

ProductionAreas <- terra::mask(ProductionAreas, DK)

cls <- data.frame(id=0:1, Protection=c("other",stringr::str_wrap("Production areas", 10)))
levels(ProductionAreas) <- cls
```

Finally, we save the production / compromised areas as a Cloud Optimised GeoTIFF:

```{r SaveProductionareas, eval=FALSE}
write_cog(ProductionAreas, "FinalLayers/ProductionAreas.tif")
```

This layer underpins the **orange category** in Fig. 1.

## Insufficient legal protection

Areas under protection schemes that do not meet the criteria for long-term legal protection (Criterion C1) or other key criteria (C2–C5) are classified as having **insufficient legal protection**.

On land, this primarily concerns **private Article 3 areas**, where:

* long-term protection is not secured, and
* there are no management requirements preventing habitat changes that would undermine biodiversity values (e.g. succession from open habitats to forest).

To operationalise this, we identify areas that:

1. are covered by **exactly one** protection scheme,
2. fall under the **Article 3** protection scheme, and
3. are on **privately owned** land.

Pixels meeting all three conditions form the “insufficient legal protection” category (yellow in Fig. 1).

### Identifying private Article 3 areas with a single protection scheme

First, we read a raster that records how many protection schemes apply to each pixel:

```{r ILP_NumberOfProtections, eval = FALSE}
NumberOfProtections <- terra::rast("Data/Rast_NumberOfProtections_Croped.tif")

# Keep only pixels with exactly one protection scheme; reclassify others to 0
NumberOfProtections <- terra::ifel(NumberOfProtections != 1, 0, NumberOfProtections)
```

Next, we read the Article 3 protection raster. In our case this layer is stored on a shared drive; the path should be adapted to your local setup. We convert it to a binary layer where 1 indicates Article 3 protection and 0 indicates “unknown” or lakes.

```{r ILP_P3, eval = FALSE}
P3 <- terra::rast("Data/Rast_p3_Croped.tif")
P3 <- as.numeric(P3)

# Remove both unknown (ukendt) and lakes (Soer):
# 0–3, 5  = Article 3 categories kept as 1
# 4, 6    = set to 0 (e.g. lakes, unknown)
P3 <- terra::ifel(P3 %in% c(0:3, 5), 1, P3)
P3 <- terra::ifel(P3 %in% c(4, 6), 0, P3)
P3 <- terra::ifel(is.na(P3), 0, P3)
```

We then read in the ownership data, convert it to a binary “private vs. other” layer, and resample it to match the resolution and extent of the Article 3 raster:

```{r ILP_Ownership, eval = FALSE}
Ownership <- terra::rast("Data/Ownership2025.tif")
Ownership <- as.numeric(Ownership)

# Keep only private areas (Ownership == 1)
Private <- Ownership
Private <- terra::ifel(Private == 1, 1, 0)

# Align to the P3 raster
Private <- terra::resample(Private, P3, method = "near")
```

Finally, we multiply the three binary layers to obtain areas that:

* are only protected by a single scheme (NumberOfProtections == 1),
* are Article 3 areas (P3 == 1), and
* are privately owned (Private == 1).

These pixels are then masked to the Denmark polygon and labelled:

```{r ILP_Combine, eval = FALSE}
Private_P3 <- Private * P3 * NumberOfProtections

# Consider only pixels within Denmark
Private_P3 <- terra::mask(Private_P3, DK)

cls <- data.frame(
  id         = 0:1,
  Protection = c("other", stringr::str_wrap("Insufficient legal protection", 10))
)
levels(Private_P3) <- cls
```

We save this layer as a Cloud Optimised GeoTIFF:

```{r ILP_Save, eval = FALSE}
write_cog(Private_P3, "FinalLayers/InsufficientLegalProtection.tif")
```

This layer corresponds to the **yellow category** (“Insufficient legal protection”) in Fig. 1 of the manuscript.


## Requires individual assessment

Some protection schemes cannot be fully evaluated with the available information (for example, because management prescriptions or long-term legal guarantees are not consistently documented at the ecosystem level). For these cases, the criteria C1–C5 cannot be conclusively assessed.

In the manuscript, areas that **require individual assessment** primarily include:

* Paragraph 3 areas (including dunes),
* Natura 2000 areas,
* conservation orders (*fredninger*, approximated here via IUCN-coded areas),
* private untouched forest,
* subsidy schemes, and
* nature and wildlife reserves.

Pixels that fall inside at least one of these schemes, and that are not already classified as fully contributing protected areas or production areas, are placed in the “requires individual assessment” category (lavender in Fig. 1).

### Identifying areas that require individual assessment

We start by reading the Paragraph 3 (including dunes) raster and converting it to a simple presence/absence layer:

```{r RIA_P3, eval = TRUE}
# Paragraph 3 and dunes
RIA_P3 <- terra::rast("Data/Rast_p3_Croped.tif")
RIA_P3 <- as.numeric(RIA_P3)
RIA_P3[!is.na(RIA_P3)] <- 1   # all non-NA treated as Paragraph 3 presence
```

Next, we read the additional protection schemes using the `ToOne()` helper defined above, which reads a raster from the `Data/` folder and converts all non-NA values to 1:

```{r RIA_layers, eval = TRUE}
Natura2000              <- ToOne("Rast_Natura2000_Croped.tif")
NaturaOgVildtreservater <- ToOne("Rast_NaturaOgVildtreservater_Croped.tif")
IUCN                    <- ToOne("Rast_IUCN_Croped.tif")
Stoette                 <- ToOne("Rast_stoette_Croped.tif")   # subsidy schemes
```

We then derive a binary layer for **private untouched forest**. In this dataset, `Rast_Urort_Skov_Croped.tif` contains information on untouched forest, and the recoding below isolates private untouched forest as a separate binary layer:

```{r RIA_privateUntouched, eval = TRUE}
PrivateUrortSkov <- terra::rast("Data/Rast_Urort_Skov_Croped.tif")
PrivateUrortSkov <- as.numeric(PrivateUrortSkov)

# Recode to a binary private untouched forest layer
PrivateUrortSkov[PrivateUrortSkov == 1] <- NA
PrivateUrortSkov[PrivateUrortSkov == 0] <- 1
PrivateUrortSkov[is.na(PrivateUrortSkov)] <- 0
```

We then sum all these layers:

* Natura 2000
* Nature and wildlife reserves
* IUCN-coded conservation orders
* Paragraph 3 and dunes
* subsidy schemes
* private untouched forest

Any pixel with a sum > 0 is considered part of the **requires individual assessment** category.

```{r RIA_combine, eval = TRUE}
RIA <- Natura2000 +
       NaturaOgVildtreservater +
       IUCN +
       RIA_P3 +
       Stoette +
       PrivateUrortSkov

RIA <- terra::mask(RIA, DK)
RIA[RIA > 0] <- 1

cls_RIA <- data.frame(
  id         = 0:1,
  Protection = c("other", stringr::str_wrap("Requires individual assessment", 10))
)
levels(RIA) <- cls_RIA
```

Finally, we save this layer as a Cloud Optimised GeoTIFF:

```{r RIA_save, eval = TRUE}
write_cog(RIA, "FinalLayers/RequiresIndividualAssessment.tif")
```

This binary raster (`RIA == 1`) corresponds to the **lavender category** (“Requires individual assessment”) in Fig. 1 of the manuscript.
